---
title: "Proximity Labelling - Mutants vs WT Student t-test script"
output: pdf_document
date: '2022-07-27'
---

```{r setup}
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(data.table)
library(tidyr)
```

# 1. Turbo

## 1.1 Data Preprocessing

```{r}
turbo_id_preprocess_all <- function(raw_data, lfq=FALSE) {
  
  if (lfq==TRUE){
    
    preprocessed <- raw_data[,c("T..Majority.protein.IDs",
                                "T..Gene.names",
                                colnames(raw_data)[grepl("LFQ.intensity.T_", 
                                                         colnames(raw_data))])]
    
  } else {
    
    preprocessed <- raw_data[,c("T..Majority.protein.IDs",
                                "T..Gene.names",
                                colnames(raw_data)[grepl("Intensity.T_", 
                                                         colnames(raw_data))])]
  
  }
  
  
  preprocessed$PROTID_temp <- sapply(preprocessed$T..Majority.protein.IDs, 
                                     function(protid) {
    unlist(strsplit(x = protid, split = ";"))[1]
    })
  preprocessed$PROTID <- gsub(preprocessed$PROTID_temp, 
                              pattern = "CON__", 
                              replacement = "")
  preprocessed$PROTID <- gsub(preprocessed$PROTID, 
                              pattern = "REV__", 
                              replacement = "")
  
  preprocessed$PROTID_temp <- NULL
  preprocessed$T..Majority.protein.IDs <- NULL
  
  preprocessed$Gene.name <- preprocessed$T..Gene.names
  preprocessed$T..Gene.names <- NULL
  
  preprocessed <- 
    preprocessed[,c("PROTID",
                    "Gene.name",
                    colnames(preprocessed)[1:(dim(preprocessed)[2]-2)])]
  
  return(preprocessed)
}
```


Comparisons between CFTR mutants and WT condition were performed based on the LFQ (Label-free quantification) log2-transformed data.


```{r}
turbo_lfq_log_pl <- read.table("raw_data/Turbo_LFQintensities_log2_export.txt",
                               sep = "\t",
                               header = T)
turbo_lfq_log_pl_df <- turbo_id_preprocess_all(turbo_lfq_log_pl,
                                               lfq=TRUE)

colnames(turbo_lfq_log_pl_df)[-1] <- gsub(x = colnames(turbo_lfq_log_pl_df)[-1],
                                          pattern = "LFQ.intensity.T_",
                                          replacement = "")
```

## 1.2 Keep proteins that are present in all replicates of all conditions.

```{r}
WT_samples <- which(grepl(pattern = "WT", colnames(turbo_lfq_log_pl_df)))
G551D_samples <- which(grepl(pattern = "G551D", colnames(turbo_lfq_log_pl_df)))
W1282X_samples <- which(grepl(pattern = "W1282X", colnames(turbo_lfq_log_pl_df)))

turbo_lfq_log_pl_df$filter <- apply(X = turbo_lfq_log_pl_df[,c(WT_samples,
                                                               G551D_samples, 
                                                               W1282X_samples)],
    MARGIN = 1,
    function(x){return(any(is.na(x)))})

turbo_filtered <- turbo_lfq_log_pl_df[which(!turbo_lfq_log_pl_df$filter),]
turbo_filtered$filter <- NULL
```

## 1.3  Comparison G551D vs WT.

### 1.3.1 Student T-test for G551D vs WT.

```{r}
turbo_filtered$t_test.G551D.pval <- 
  sapply(1:nrow(turbo_filtered), function(iprot){
    return(t.test(as.numeric(turbo_filtered[iprot,WT_samples]),
                  as.numeric(turbo_filtered[iprot,G551D_samples]))$p.value)
    })
# Correction for multiple testing
turbo_filtered$t_test.G551D.padj <- p.adjust(turbo_filtered$t_test.G551D.pval,
                                             method = "BH")

#-log10(p.value) for the plot
turbo_filtered$t_test.G551D.log.pval <- -log10(turbo_filtered$t_test.G551D.pval)

# log2(FoldChange)
turbo_filtered$diff.mean.G551D.WT <- sapply(1:nrow(turbo_filtered), 
                                            function(iprot){
  return(mean(as.numeric(turbo_filtered[iprot,G551D_samples])) -
           mean(as.numeric(turbo_filtered[iprot,WT_samples])))
  })

# Over- or under-expressed class for the volcano plot
turbo_filtered$G551D.class <- sapply(1:nrow(turbo_filtered), function(i_row){
  
  p.val <- turbo_filtered[i_row, "t_test.G551D.pval"]
  FC <- turbo_filtered[i_row, "diff.mean.G551D.WT"]
  
  if (p.val<0.1 & FC<0){
    return("DOWN")
  } else if (p.val<0.1 & FC>0){
    return("UP")
  } else {
    return("NO")
  }
})
turbo_filtered$G551D.class <- factor(turbo_filtered$G551D.class,
                                     levels = c("NO", "DOWN", "UP"))
```

### 1.3.2 Volcano plot for G551D vs WT.

```{r}
volcano_turbo_G551D <- ggplot(turbo_filtered, aes(x = diff.mean.G551D.WT, 
                                                  y = t_test.G551D.log.pval, 
                                                  color = G551D.class, 
                                                  fill = G551D.class))+
  geom_point(pch=21, size=0.5, stroke=0.2)+
  ylim(0, 5)+
  scale_color_manual(values=c("#999999","#737373", "#737373"))+
  scale_fill_manual(values=c("#bfbfbf","#ed5e5f", "#69a3d2"))+
  theme_bw()+
  theme(legend.position="none",
        text=element_text(size = 12),
        title = element_text(size = 12))+
  ggtitle("TurboID - G551D") +
  guides(fill="legend")+
  scale_x_continuous(breaks=c(-4, -2, 0, 2, 4), limits=c(-4.5,4.5))+
  labs(y=expression(-Log[10]~p-value),
       x=expression(Log[2]~FC))

volcano_turbo_G551D
```

## 1.4  Comparison W1282X vs WT.

### 1.4.1 Student T-test for W1282X vs WT.

```{r}
turbo_filtered$t_test.W1282X.pval <- sapply(1:nrow(turbo_filtered), function(iprot){
  return(t.test(as.numeric(turbo_filtered[iprot,WT_samples]),
                as.numeric(turbo_filtered[iprot,W1282X_samples]))$p.value)
})

# Correction for multiple testing
turbo_filtered$t_test.W1282X.padj <- p.adjust(turbo_filtered$t_test.W1282X.pval,
                                             method = "BH")

#-log10(p.value) for the plot
turbo_filtered$t_test.W1282X.log.pval <- -log10(turbo_filtered$t_test.W1282X.pval)

# log2(FoldChange)
turbo_filtered$diff.mean.W1282X.WT <- sapply(1:nrow(turbo_filtered), 
                                             function(iprot){
  return(mean(as.numeric(turbo_filtered[iprot,W1282X_samples]))-
           mean(as.numeric(turbo_filtered[iprot,WT_samples])))})

# Over- or under-expressed class for plot
turbo_filtered$W1282X.class <- sapply(1:nrow(turbo_filtered), function(i_row){
  
  p.val <- turbo_filtered[i_row, "t_test.W1282X.pval"]
  FC <- turbo_filtered[i_row, "diff.mean.W1282X.WT"]
  
  if (p.val<0.1 & FC<0){
    return("DOWN")
  } else if (p.val<0.1 & FC>0){
    return("UP")
  } else {
    return("NO")
  }
})
turbo_filtered$W1282X.class <- factor(turbo_filtered$W1282X.class,
                                     levels = c("NO", "DOWN", "UP"))
```

### 1.4.2 Volcano plot for W1282X vs WT.

```{r, warning=FALSE}
volcano_turbo_W1282X <- ggplot(turbo_filtered, aes(x = diff.mean.W1282X.WT, 
                                                   y = t_test.W1282X.log.pval, 
                                                   fill = W1282X.class,
                                                   color = W1282X.class))+
  geom_point(shape=21, size=0.5, stroke=0.2)+
  ylim(0, 5)+
  scale_color_manual(values=c("#999999","#737373", "#737373"))+
  scale_fill_manual(values=c("#bfbfbf","#ed5e5f", "#69a3d2"))+
  theme_bw()+
  theme(legend.position="none",
        text=element_text(size = 12),
        plot.title = element_text(size = 12),
        axis.title = element_text(size=12))+
  ggtitle("TurboID - W1282X") +
  guides(fill="legend")+
  scale_x_continuous(breaks=c(-4, -2, 0, 2, 4), limits=c(-4.5,4.5))+
  labs(y=expression(-Log[10]~p-value),
       x=expression(Log[2]~FC))

volcano_turbo_W1282X
```

# 2. APEX

## 2.1 Data Preprocessing

```{r}
apex_preprocess_all <- function(raw_data, lfq=FALSE) {
  
  if (lfq==TRUE){
    
    preprocessed <- raw_data[,c("T..Majority.protein.IDs",
                                "T..Gene.names",
                                colnames(raw_data)[grepl("LFQ.intensity.A_", 
                                                         colnames(raw_data))])]
    
  } else {
    
    preprocessed <- raw_data[,c("T..Majority.protein.IDs",
                                "T..Gene.names",
                                colnames(raw_data)[grepl("Intensity.A_", 
                                                         colnames(raw_data))])]
  }
  
  preprocessed$PROTID_temp <- sapply(preprocessed$T..Majority.protein.IDs, 
                                     function(protid) {
                                       unlist(strsplit(x = protid, 
                                                       split = ";"))[1]})
  preprocessed$PROTID <- gsub(preprocessed$PROTID_temp, 
                              pattern = "CON__", 
                              replacement = "")
  preprocessed$PROTID <- gsub(preprocessed$PROTID, 
                              pattern = "REV__", 
                              replacement = "")
  
  preprocessed$PROTID_temp <- NULL
  preprocessed$T..Majority.protein.IDs <- NULL
  
  preprocessed$Gene.name <- preprocessed$T..Gene.names
  preprocessed$T..Gene.names <- NULL
  
  preprocessed <- 
    preprocessed[,c("PROTID",
                    "Gene.name",
                    colnames(preprocessed)[1:(dim(preprocessed)[2]-2)])]
  
  return(preprocessed)
  
}
```


Comparisons between CFTR mutants and WT condition were performed based on the LFQ (Label-free quantification) log2-transformed data.


```{r}
apex_lfq_log_pl <- read.table("raw_data/Apex_LFQintensities_log2_export.txt",
                               sep = "\t",
                               header = T)
apex_lfq_log_pl_df <- apex_preprocess_all(apex_lfq_log_pl,
                                               lfq=TRUE)

colnames(apex_lfq_log_pl_df)[-1] <- gsub(x = colnames(apex_lfq_log_pl_df)[-1],
                                          pattern = "LFQ.intensity.A_",
                                          replacement = "")
```

## 2.2 Keep proteins that are present in all replicates of all conditions.

```{r}
WT_samples <- which(grepl(pattern = "WT", colnames(apex_lfq_log_pl_df)))
G551D_samples <- which(grepl(pattern = "G551D", colnames(apex_lfq_log_pl_df)))
W1282X_samples <- which(grepl(pattern = "W1282X", colnames(apex_lfq_log_pl_df)))

apex_lfq_log_pl_df$filter <- apply(X = apex_lfq_log_pl_df[,c(WT_samples,
                                                             G551D_samples, 
                                                             W1282X_samples)],
                                    MARGIN = 1,
                                    function(x){return(any(is.na(x)))})

apex_filtered <- apex_lfq_log_pl_df[which(!apex_lfq_log_pl_df$filter),]
apex_filtered$filter <- NULL
```

## 2.3  Comparison G551D vs WT.

### 2.3.1 Student T-test for G551D vs WT.

```{r}
apex_filtered$t_test.G551D.pval <- sapply(1:nrow(apex_filtered), function(iprot){
  return(t.test(apex_filtered[iprot,WT_samples],
                apex_filtered[iprot,G551D_samples])$p.value)
})
# Correction for multiple testing
apex_filtered$t_test.G551D.padj <- p.adjust(apex_filtered$t_test.G551D.pval,
                                             method = "BH")

#-log10(p.value) for the plot
apex_filtered$t_test.G551D.log.pval <- -log10(apex_filtered$t_test.G551D.pval)

# log2(FoldChange)
apex_filtered$diff.mean.G551D.WT <- sapply(1:nrow(apex_filtered), function(iprot){
  return(mean(as.numeric(apex_filtered[iprot,G551D_samples])) -
           mean(as.numeric(apex_filtered[iprot,WT_samples])))})

# Over- or under-expressed class for the volcano plot
apex_filtered$G551D.class <- sapply(1:nrow(apex_filtered), function(i_row){
  
  p.val <- apex_filtered[i_row, "t_test.G551D.pval"]
  FC <- apex_filtered[i_row, "diff.mean.G551D.WT"]
  
  if (p.val<0.1 & FC<0){
    return("DOWN")
  } else if (p.val<0.1 & FC>0){
    return("UP")
  } else {
    return("NO")
  }
})
apex_filtered$G551D.class <- factor(apex_filtered$G551D.class,
                                     levels = c("NO", "DOWN", "UP"))
```

### 2.3.2 Volcano plot for G551D vs WT.

```{r}
volcano_apex_G551D <- ggplot(apex_filtered, aes(x = diff.mean.G551D.WT, 
                                                y = t_test.G551D.log.pval, 
                                                color = G551D.class, 
                                                fill = G551D.class))+
  geom_point(pch=21, size=0.5, stroke=0.2)+
  ylim(0, 5)+
  scale_color_manual(values=c("#999999","#737373", "#737373"))+
  scale_fill_manual(values=c("#bfbfbf","#ed5e5f", "#69a3d2"))+
  theme_bw()+
  theme(legend.position="none",
        text=element_text(size = 12),
        title = element_text(size = 12))+
  ggtitle("APEX - G551D") +
  guides(fill="legend")+
  scale_x_continuous(breaks=c(-4, -2, 0, 2, 4), limits=c(-4.5,4.5))+
  labs(y=expression(-Log[10]~p-value),
       x=expression(Log[2]~FC))

volcano_apex_G551D
```

## 2.4  Comparison W1282X vs WT.

### 2.4.1 Student T-test for W1282X vs WT.

```{r}
apex_filtered$t_test.W1282X.pval <- sapply(1:nrow(apex_filtered), function(iprot){
  return(t.test(apex_filtered[iprot,WT_samples],
                apex_filtered[iprot,W1282X_samples])$p.value)
})

# Correction for multiple testing
apex_filtered$t_test.W1282X.padj <- p.adjust(apex_filtered$t_test.W1282X.pval,
                                               method = "BH")

#-log10(p.value) for the plot
apex_filtered$t_test.W1282X.log.pval <- -log10(apex_filtered$t_test.W1282X.pval)

# log2(FoldChange)
apex_filtered$diff.mean.W1282X.WT <- sapply(1:nrow(apex_filtered), function(iprot){
  return(mean(as.numeric(apex_filtered[iprot,W1282X_samples])) -
           mean(as.numeric(apex_filtered[iprot,WT_samples])))})

# Over- or under-expressed class for plot
apex_filtered$W1282X.class <- sapply(1:nrow(apex_filtered), function(i_row){
  
  p.val <- apex_filtered[i_row, "t_test.W1282X.pval"]
  FC <- apex_filtered[i_row, "diff.mean.W1282X.WT"]
  
  if (p.val<0.1 & FC<0){
    return("DOWN")
  } else if (p.val<0.1 & FC>0){
    return("UP")
  } else {
    return("NO")
  }
})
apex_filtered$W1282X.class <- factor(apex_filtered$W1282X.class,
                                      levels = c("NO", "DOWN", "UP"))
```

### 2.4.2 Volcano plot for W1282X vs WT.

```{r}
volcano_apex_W1282X <- ggplot(apex_filtered, aes(x = diff.mean.W1282X.WT, 
                                                 y = t_test.W1282X.log.pval, 
                                                 color = W1282X.class, 
                                                 fill = W1282X.class))+
  geom_point(pch=21, size=0.5, stroke=0.2)+
  ylim(0, 5)+
  scale_color_manual(values=c("#999999","#737373", "#737373"))+
  scale_fill_manual(values=c("#bfbfbf","#ed5e5f", "#69a3d2"))+
  theme_bw()+
  theme(legend.position="none",
        text=element_text(size = 12),
        title = element_text(size = 12))+
  ggtitle("APEX - W1282X") +
  guides(fill="legend")+
  scale_x_continuous(breaks=c(-4, -2, 0, 2, 4), limits=c(-4.5,4.5))+
  labs(y=expression(-Log[10]~p-value),
       x=expression(Log[2]~FC))

volcano_apex_W1282X
```